Collective excitations in competing phases in two and three dimensions

### 2024-03-27 paper/main.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                  Abstract                                                 %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
    We investigate the superconducting (SC), charge-density wave (CDW), and antiferromagnetic (AFM) phases in the 
		extended Hubbard model at zero temperature and half-filling. 
    We employ the iterated equations of motion approach to compute the two-particle Green's functions and their
		spectral densities. 
    This renders a comprehensive analysis of the behavior of collective excitations possible as the 
		model's parameters are tuned across phase transitions. 
    We identify the well-known amplitude (Higgs) and phase (Anderson-Bogoliubov) modes within the 
		superconducting phase and observe a similar excitation (cooperon) in the CDW phase which shifts towards 
		zero energy as the system approaches the phase transition to the SC phase. 
    In the CDW phase, close to the phase transition to the AFM phase, 
    we find a collective mode, an exciton, that does not change significantly 
    and another mode, a longitudinal magnon, that emerges from the two-particle continuum as the system approaches the phase transition to the AFM phase. 
    \red{Habe ich geaendert - stimmt es so?} \blue{Nicht ganz, das Magnon verschwindet, wenn man weg von der Phasengrenze geht, also zu kleineren U/größeren V. 
    Umgekehrt in der AFM phase, verschwindet das Exziton, bei dem analogen Prozess, also zu großem U/kleinen V. Des Weiteren sind beide Magnonen außerhalb der AFM Phase identisch.
    Ist es dann richtig, hier nur von dem longitudinalen zu sprechen?}
		In the AFM phase, we identify both a longitudinal (amplitude) and a 
		transversal Goldstone magnon; the latter is located at zero energy.
\end{abstract}

@@

On the theoretical side, we choose a paradigmatic, but simple model which displays competing phases.
The Hubbard model has been employed in a plethora of previous studies and competing phases are established in its extensions. 
Early studies proved the existence of eigenstates of the Hubbard Hamiltonian that exhibit off-diagonal long-range order, encouraging the model's usage for the description of high-temperature superconductivity \cite{yang89}.
Shortly afterward, an exact SO(4) symmetry was discovered, which induces a degeneracy of 
superconductivity (SC) and a charge-density wave (CDW) governed by an attractive on-site interaction \cite{yang90}.
This coexistence stems from the existence of a particle-hole transformation on bipartite lattices, 
that maps the attractive Hubbard model rigorously onto the repulsive one, 
exhibiting antiferromagnetism (AFM) \cite{Hirsch85}.
The order parameters of the previously mentioned SC and CDW phases map to different spin expectation values \cite{zitko15,lieb89}.
\blue{Den vorigen Satz verstehe ich leider gar nicht: phase = spin operator? Ja, das war ungenau. So ist es vermutlich besser?}

@@

Moreover, we demonstrate an explicit way to compute various Green's functions by this approach.
By extension, the corresponding spectral functions of the investigated systems can be computed and 
we discuss the signatures of collective excitations in the spectral functions. \blue{within these?}
The most prominent examples in the SC phase are the well-known phase mode (Anderson-Bogoliubov) and 
the amplitude (Higgs) modes. The former occurs in neutral superfluids at zero energy and was found
in a large number of studies \cite{Bogoljubov1958,Anderson58,Brieskorn74,Schmid1975,simanek1975,schon76,Maiti2015,Sun2020,Fan22} which we cannot list exhaustively here.
We obtain this mode based on a microscopic description without long-range electromagnetic interactions.
The inclusion of this kind of interaction would shift this mode towards the plasma frequency 
\cite{Anderson58,schon76,Kulik1981} which is, however, beyond the scope of the present paper. 
The amplitude mode, on the other hand, is located at the lower edge of the quasiparticle continuum.
The corresponding energy is the energy required to break up a Cooper pair 
\cite{Schmid1975,Varma02,Cea14,Measson14,Tsuji15,Krull16,Mueller2019,Schwarz20}.
This Higgs mode is not charged so that it does not couple to the electromagnetic fields.

@@

If $\check{N}_X$ has a singular part, for instance, due to numerical inaccuracies \blue{Das kann auch völlig analytisch auftreten}, the inverse can be replaced by the Moore-Penrose pseudo-inverse. 
A possible way to compute this inverse is to diagonalize 
$\check{N}_X = \mathcal{V} \mathcal{D} \mathcal{V}^\dagger$ yielding a diagonal matrix $\mathcal{D}$ and a unitary transformation matrix $\mathcal{V}$. 
Then, the pseudo-inverse can be defined by
\begin{equation}
    \check{N}^{-1} \coloneqq \mathcal{V} \mathcal{D}^{-1} \mathcal{V}^\dagger
\end{equation}
with \red{Ich habe das hoch + in ein hoch -1 geaendert, weil mir das "`+"' zu sehr an Transposition erinnerte.}
\begin{equation}
    (\mathcal{D}^{-1})_{ij} = \begin{cases}
        1/\mathcal{D}_{ij} & \mathcal{D}_{ij} \neq 0 \\ 0 & \text{otherwise}
    \end{cases}.
\end{equation}
Note that this definition recovers the standard inverse for invertible matrices.

@@

Next, we analyze the peak in $\spectral{CDW}$ occurring if the system is in the SC phase for $V<0$.
A plot of its position $\omega_0$ and weight $W_0$ is given in Fig.\ \ref{fig:cdw_in_sc_behavior}.
The upper panels (a) and (b) depict the peak position and weight, respectively.
For $|V| \ll t$, the behavior of both of them is the same as for the 
previously discussed peak in $\spectral{SC}$ in the CDW phase.
Increasing $|V|$, the increase of the peak position slows down while the weights start to drop rapidly.
We cannot compute the Green's functions for even larger $|V|$ with $V<0$ because the dynamical matrix $\mM$ 
acquires negative eigenvalues below $V\approx -0.28t$ (square lattice) or $V\approx -0.34t$ (simple cubic lattice). 
We recall that this indicates that the assumed phase, i.e., the SC phase is not stable. 
Therefore, we conclude that for sufficiently negative values of $V$, 
the superconducting phase is not the true ground state as indicated by the red shading in Fig.\ \ref{fig:phase_diagram}.
Specifically, the region of the SC phase on the square lattice 
appears to be even smaller than shown by the $d_{x^2 -y^2}$ data from Ref.\ \cite{Micnas88b}.
Recently, there has been evidence for phase separation for negative values of $V$ 
on the square lattice, at least at finite temperature \cite{Linner23}. 
Thus, we interpret the negative eigenvalues in the dynamical matrix as indicators of phase separation. 
Physically, a strong attractive interaction clearly favors phase separation since the particles prefer to stay close together 
so it is highly plausible that they gather in one region of the sample while leaving the remainder essentially empty.
\red{Was ist in 3D - gibt es da nicht auch negative Eigenwerte usw.? Das solten wir dann hier
ebenfalls besprechen. Wo wollten wir denn die Diskussion des Tricks von Biermann publizieren?}

@@

Last, we investigate how the modes behave as the system passes from the CDW to the AFM phase.
As in the previous subsection, we plot the occurring peaks in $\spectral{CDW}$ and 
$\spectral{l.AFM}$ in the respective phases in Fig.\ \ref{fig:afm_cdw_peaks_overview}.
On the right-hand side of each panel, i.e., for $U > zV$, the system is in the AFM phase (green shading);
on the left-hand side, it is in the CDW phase (orange shading).
Qualitatively, the peak in $\spectral{l.AFM}$ behaves like the peak in 
$\spectral{CDW}$ if the phases are swapped. \red{Ist das exakt oder nur ungefaehr so?} \blue{tatsächlich exakt.}

@@

Lastly, all other peaks below the two-particle continuum behave identically.
The real part of them behaves like $1/\omega$ in close vicinity to the peak position.
This indicates that the peaks are $\delta$ distributions.
The peaks of $\spectral{l.AFM}$ and $\spectral{CDW}$ close to the phase transition are identical 
when swapped, i.e., $\spectral{l.AFM}$ in the CDW phase is the same as $\spectral{CDW}$ in the AFM phase.
This reflects the SO(4) symmetry of the model. \red{Die Spektralfunktionen sind also komplett identisch, richtig?
Oder nur so ungefähr?} \blue{Wirklich komplett identisch}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                  Abstract                                                 %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
    We investigate the superconducting (SC), charge-density wave (CDW), and antiferromagnetic (AFM) phases in the 
		extended Hubbard model at zero temperature and half-filling. 
    We employ the iterated equations of motion approach to compute the two-particle Green's functions and their
		spectral densities. 
    This renders a comprehensive analysis of the behavior of collective excitations possible as the 
		model's parameters are tuned across phase transitions. 
    We identify the well-known amplitude (Higgs) and phase (Anderson-Bogoliubov) modes within the 
		superconducting phase and observe a similar excitation (cooperon) in the CDW phase which shifts towards 
		zero energy as the system approaches the phase transition to the SC phase. 
    In the CDW phase, close to the phase transition to the AFM phase, 
    we find a collective mode, an exciton, that does not change significantly 
    and another mode, a longitudinal magnon, 
    that emerges from the two-particle continuum as the system approaches the phase transition to the AFM phase. 
    It becomes identical with the former at the transition.
		In the AFM phase, their roles are reversed.
    Additionally, we find a transversal Goldstone magnon located at zero energy.
\end{abstract}

@@

On the theoretical side, we choose a paradigmatic, but simple model which displays competing phases.
The Hubbard model has been employed in a plethora of previous studies and competing phases are established in its extensions. 
Early studies proved the existence of eigenstates of the Hubbard Hamiltonian that exhibit off-diagonal long-range order, encouraging the model's usage for the description of high-temperature superconductivity \cite{yang89}.
Shortly afterward, an exact SO(4) symmetry was discovered, which induces a degeneracy of 
superconductivity (SC) and a charge-density wave (CDW) governed by an attractive on-site interaction \cite{yang90}.
This coexistence stems from the existence of a particle-hole transformation on bipartite lattices, 
that maps the attractive Hubbard model rigorously onto the repulsive one, 
exhibiting antiferromagnetism (AFM) \cite{Hirsch85}.
The order parameters of the SC and CDW phases in the attractive model map to different spin expectation values in the repulsive one \cite{zitko15,lieb89}.

@@

Moreover, we demonstrate an explicit way to compute various Green's functions by this approach.
By extension, the corresponding spectral functions of the investigated systems can be computed and 
we discuss the signatures of collective excitations in the spectral functions. 
The most prominent examples in the SC phase are the well-known phase mode (Anderson-Bogoliubov) and 
the amplitude (Higgs) modes. The former occurs in neutral superfluids at zero energy and was found
in a large number of studies \cite{Bogoljubov1958,Anderson58,Brieskorn74,Schmid1975,simanek1975,schon76,Maiti2015,Sun2020,Fan22} which we cannot list exhaustively here.
We obtain this mode based on a microscopic description without long-range electromagnetic interactions.
The inclusion of this kind of interaction would shift this mode towards the plasma frequency 
\cite{Anderson58,schon76,Kulik1981} which is, however, beyond the scope of the present paper. 
The amplitude mode, on the other hand, is located at the lower edge of the quasiparticle continuum.
The corresponding energy is the energy required to break up a Cooper pair 
\cite{Schmid1975,Varma02,Cea14,Measson14,Tsuji15,Krull16,Mueller2019,Schwarz20}.
This Higgs mode is not charged so that it does not couple to the electromagnetic fields.

@@

If $\check{N}_X$ has a singular part, for instance, due to numerical inaccuracies or for particular operators, the inverse can be replaced by the Moore-Penrose pseudo-inverse. 
A possible way to compute this inverse is to diagonalize 
$\check{N}_X = \mathcal{V} \mathcal{D} \mathcal{V}^\dagger$ yielding a diagonal matrix $\mathcal{D}$ and a unitary transformation matrix $\mathcal{V}$. 
Then, the pseudo-inverse can be defined by
\begin{equation}
    \check{N}^{-1} \coloneqq \mathcal{V} \mathcal{D}^{-1} \mathcal{V}^\dagger
\end{equation}
with \red{Ich habe das hoch + in ein hoch -1 geaendert, weil mir das "`+"' zu sehr an Transposition erinnerte.}
\begin{equation}
    (\mathcal{D}^{-1})_{ij} = \begin{cases}
        1/\mathcal{D}_{ij} & \mathcal{D}_{ij} \neq 0 \\ 0 & \text{otherwise}
    \end{cases}.
\end{equation}
Note that this definition recovers the standard inverse for invertible matrices.

@@

Next, we analyze the peak in $\spectral{CDW}$ occurring if the system is in the SC phase for $V<0$.
A plot of its position $\omega_0$ and weight $W_0$ is given in Fig.\ \ref{fig:cdw_in_sc_behavior}.
The upper panels (a) and (b) depict the peak position and weight, respectively.
For $|V| \ll t$, the behavior of both of them is the same as for the 
previously discussed peak in $\spectral{SC}$ in the CDW phase.
Increasing $|V|$, the increase of the peak position slows down while the weights start to drop rapidly.
We cannot compute the Green's functions for even larger $|V|$ with $V<0$ because the dynamical matrix $\mM$ 
acquires negative eigenvalues below $V\approx -0.28t$ (square lattice) or $V\approx -0.34t$ (simple cubic lattice). 
We recall that this indicates that the assumed phase, i.e., the SC phase is not stable. 
Therefore, we conclude that for sufficiently negative values of $V$, 
the superconducting phase is not the true ground state as indicated by the red shading in Fig.\ \ref{fig:phase_diagram}.
Specifically, the region of the SC phase on the square lattice 
appears to be even smaller than shown by the $d_{x^2 -y^2}$ data from Ref.\ \cite{Micnas88b}.
Recently, there has been evidence for phase separation for negative values of $V$ 
on the square lattice, at least at finite temperature \cite{Linner23}. 
Thus, we interpret the negative eigenvalues in the dynamical matrix as indicators of phase separation. 
Physically, a strong attractive interaction clearly favors phase separation since the particles prefer to stay close together 
so it is highly plausible that they gather in one region of the sample while leaving the remainder essentially empty.
In 3D, we observe negative eigenvalues in $\mM$ for strongly negative interactions as well.
Thus, we expect that phase separation also occurs on this lattice.

@@

Last, we investigate how the modes behave as the system passes from the CDW to the AFM phase.
As in the previous subsection, we plot the occurring peaks in $\spectral{CDW}$ and 
$\spectral{l.AFM}$ in the respective phases in Fig.\ \ref{fig:afm_cdw_peaks_overview}.
On the right-hand side of each panel, i.e., for $U > zV$, the system is in the AFM phase (green shading);
on the left-hand side, it is in the CDW phase (orange shading).
Qualitatively, the peak in $\spectral{l.AFM}$ behaves exactly like the peak in 
$\spectral{CDW}$ if the phases are swapped.
We attribute this behavior to the simplicity of our model and do not expect this behavior quantitatively generic.

@@

Lastly, all other peaks below the two-particle continuum behave identically.
The real part of them behaves like $1/\omega$ in close vicinity to the peak position.
This indicates that the peaks are $\delta$ distributions.
The peaks of $\spectral{l.AFM}$ and $\spectral{CDW}$ close to the phase transition are identical 
when swapped, i.e., $\spectral{l.AFM}$ in the CDW phase is the same as $\spectral{CDW}$ in the AFM phase.
However, we do not believe that this identity is quantitatively generic.
